Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language

### 2022-08-06 manuscript_v2.tex
model all vs camera only
The plot is added for the comparison of all vs camera only participants.
updated files

Sign Languages are natural languages of the Deaf communities all around
the world. They are independent languages, which do not derive from
majority spoken language(s) but emerge naturally when deaf people form a
community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has
been the officially recognized language of the Deaf community in Turkey
since 2005 (see Nuhbalaoglu, 2018 for a brief review of the historical
development of the language). Deaf or hearing children acquiring a SL
within a family where there is one or more deaf caregivers display the
same developmental milestones as hearing children acquiring a spoken
language in a hearing family (Chamberlain et al., 1999; Meier and
Newport, 1990). However, reports indicate that over 90\% of all deaf
individuals are born into hearing families who do not know how to sign
(i.e., those with language deprivation) (Woll, 2013). This means that
the auditory modality used in the household is not accessible to many
deaf children. That is why the age of acquisition of a first functional
language varies greatly among deaf children, with many of whom being
deprived of their first languages typically until after age 7 when they
start formal education in a deaf school. First language deprivation or
delay has been found to have negative effects on certain cognitive and
linguistic measures among late signing children. To name a few, delayed
exposure to a first functional language is reported to negatively affect
deaf children's morphosyntactic production and processing (Boudreault
and Mayberry, 2006; Cheng and Mayberry, 2021; Cormier et al., 2012;
Karadöller et al., 2017; Kayabaşı and Gökgöz, 2022; Newport, 1990) and
narrative abilities (e.g., Becker, 2009). In our overall research, we
are concerned with the effects of this late exposure in the ultimate
adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy (Gundel et al., 1993).
According to this model, given or familiar information is expressed in
more attenuated forms such as pronominal use and null marking since the
mental status of the expressed referent is already at the center of
attention, as indicated by the present contextual information (Chafe,
1976). The Givenness Hierarchy is akin to other previously proposed
models of reference control within the topicality and saliency paradigm
such as the Accessibility Hierarchy (Ariel, 1990), Topic Continuity
(Givon, 1983), and Familiarity Scale (Prince, 1981). Gundel et al.
(1993) argued that the distribution of forms is not random but abides by
the Gricean Maxim of Quantity (Grice, 1975). That is to say, individuals
must make the most informative and succinct choice possible to retain
economy of form. When multiple expressions are available for the same
referent as in an utterance like (\ref{ex:1}), phonetically simplest
form is selected with the least redundant information are selected based
on an economy principle. In the utterance, REs for ``Eddie'' are given
in bold. All three REs (nominal, pronominal, and null marking) refer to
the same entity; however, the quantity of marking decreases as the
referent's accessibility increases. On this ground, for highly
accessible referents (i.e., those that are highly active in the mind),
linguistic forms with the least phonetic content are selected for the
least amount of processing effort.

@@

Commonly in both spoken and sign languages nouns and noun phrases are
used as REs. Spoken languages use pronouns, determines and
demonstratives where sign languages use a pointing sign, usually glossed
as IX, which assumes similar functions (Lillo-Martin and Klima, 1990).
In addition, the pointing sign is used to referentially anchor a
referent to a location in signing space. For instance, when the signer
introduces \textsc{elvan} for the first time, s/he can assign her to the
right. This referential locus can later be used by agreement verbs and
classifier predicates (more below). For instance, in (\ref{ex:2a}) there
are two referents, \textsc{elvan}, who is referentially-assigned to the
right side of the signer, represented with the ``a'' subscript and
\textsc{demet}, who is referentially-assigned to the left side of the
signer, represented with the ``b'' subscript. When the signer makes a
contrast in (\ref{ex:2b}), as a pronominal use, she points to the right
side to refer to ELVAN and to the left to refer to \textsc{demet}. We
schematically represents the referential locus assigned to the referents
in (\ref{ex:3}).

@@

Grammatical gender as in Romance languages or noun classes as in Bantu
languages are some reference tracking tools that are familiar from
spoken languages. A somehow less known reference-tracking tool in spoken
languages is \emph{verbal classifiers}, which morphologically attach to
the verb and often cross-reference a core argument in a clause
(Aikhenvald, 2017). Classifiers denote sets of objects based on semantic
criteria such as size, shape, body-part, texture and sometimes function.
Usually, the verbal classifier and the cross-referenced noun co-occur in
the same clause when the argument is introduced into discourse. The
classifier can track the argument on its own on subsequent occurrences.
In sign languages, a verbal classifier morpheme is encoded by a
\emph{handshape} that is articulated phonologically co-temporally with a
movement, which encodes the verbal meaning (Benedicto and Brentari,
2004; Supalla, 1990). These two infused morphemes form a complex
predicate. Some classifier handshapes and the set of objects that they
denote are below. Some classifier handshapes and the set of objects that
they denote are presented below in (\ref{ex:4}).

@@

In (\ref{ex:5}), in the first clause, a referential locus, i.e.~locus-a,
is assigned to the locative source argument \textsc{cage}. In the second
clause, the core argument \textsc{bird} is cross-referenced with the
classifier handshape, i.e.~small-legged animal. It flies from the
referential locus of the \textsc{cage} (locus-a) to the referential
locus of the goal \textsc{tree} (locus-b).

@@

    \ex \textsc{tree} assigned to the referential locus-b. The source and goal tracked on the predicates \textsc{fly} and \textsc{land}. \\ \\
    \textsc{bird fly[manner]-a tree-b land. CL-b}\\
    ‘The bird flies (from the cage assigned to the left) to a tree (assigned to the right).’
    \end{xlist}
    \end{exe}
    \hfill (Adapted from Gökgöz and Sevgi 2020:318)

@@

Classifier predicates are highly complex morphological constructions
which are very common in sign languages (Supalla, 1986). These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape (Talmy, 2003). As in most sign languages hitherto studied,
classifier constructions are also common and found in TİD (see Dikyuva
et al., 2015 for a review). Following the classification of
Engberg-Pedersen (1993), we distinguish between four types of classifier
constructions. Whole entity classifiers (WECL) represent the entirety of
any referent whereas body part classifiers (BPCL) only reflect a
specific part of an animate entity (e.g., limbs, head, tongue, teeth,
and so forth). Handling classifiers (HCL) occur with transitive verbs
and depict how an object is handled or manipulated by agent. On this
ground, an HCL\footnote{As discussed before, an HCL is transitive in
  nature and depicts the handling action performed by ananimate agent.
  This is why most research tend to consider HCLs as an extension to
  constructed action (Cormier et al., 2013; Dudis, 2004).} depicts the
manner of the referring action while the former two constructions encode
the location of the entity as well as its motion in space. Extension
classifiers (ExtCL), on the other hand, denote the size and shape of an
entity by following a trajectory to trace its shape. We exemplify these
constructions in (\ref{ex:8}).

@@

Like in first language development, narrative skills have been
extensively studied among second language learners (see Bel et al.,
2015; Frederiksen and Mayberry, 2019). Comparing 11 deaf adults signers
and 13 hearing advanced learners of Catalan Sign Language (Llengua de
Signes Catalana - LSC), Bel et al. (2015) found that hearing signers
resort to more overt forms (nominals and pronouns) than native signers
to avoid potential ambiguities among referents. Nevertheless, their
results did not reach statistical significant for the interaction of
discourse status and acquisition group. A similar line of research came
from Frederiksen and Mayberry (2019) who found a statistically
significant difference between both groups, that is native signers and
hearing L2 learners of ASL, only for classifier use especially in
re-introduced contexts. L2 learners in the study used more nominals and
fewer instances of null marking in number, indicative of a limited
effect of overexplicitness among second language learners. However,
these studies treated late and native deaf signers as a single group.
That is, the control group consisted of both native signers who had one
or more deaf parents and also native-like (Cormier et al., 2012) late
deaf signers who had hearing parents and their age of acquisition
varied. Given the preliminary evidence that topic continuity and
salience marking is also difficult for second language learners
regardless of modality, it has been suggested under the Interface
Hypothesis that the syntax-discourse interface is vulnerable and
especially challenging to acquire (Sorace, 2011) for second language
learners. Similarly, Sorace and Serratrice (2009) discuss that this
interface entails much more than morphosyntactic knowledge but also
includes cognitive and pragmatic resources that help provide discourse
coherence. They propose that the second language discourse is
over-redundant. This means that second language learners produce more
overt pronouns or fuller forms of language for highly accessible
referents possibly as a compensatory strategy in an attempt to reduce
the increasing load on cognitive processing.

Still, only very few research examined whether linguistically deprived
signers (i.e., deaf signers with delayed exposure to a sign language)
also display redundancy in their reference tracking. Becker (2009)
examined 6 DGS signing deaf children (3 native and 3 late)`s narrative
abilities. She reported that late signing children generated more
explicit forms such as nominals when referring to the characters in the
story compared to native learners, concluding that first language
deprivation results in an inadequate representation of the addressee and
less efficient use of morphosyntax. Comparably, Cormier et al. (2013)
investigated the use of CA in 15 BSL among deaf native and late adult
signers and found that late signers were redundant in topic continuity.
In other words, the late signers produced more explicit forms like
nominals, and fewer instances of null marking when maintaining previous
referents. Most relevantly to the present study, Gür (2018) studied the
development of narrative skills among 10 native and 10 late deaf
children and deaf adult signers of TİD. She found that native-signing
children and adults had more instances of explicit referent
introduction, set the scene more frequently, and employed CA in an
appropriate manner more than corresponding late-signing children and
adults. Although this study presented preliminary findings on age of
acquisition effects on narrative competence, it does not investigate how
native and late signers employ different linguistic tools based on the
accessibility of the referent. Therefore, to our knowledge, there is a
lack of research directly examining the effects of first language
deprivation on deaf adult signers' selection of REs in signed
narratives.

@@

Overall, considering what has been suggested for the distribution of REs
in signed narratives, we expect more explicit forms like nominals for
referent introduction and and less explicit forms such as null marking,
pronominal and/or classifier use for referent maintainance. For the
re-introduction contexts, where referents which are discontinued across
one or more clauses are brought back into the discourse, we expect a
less clear picture. Some re-introduced referents will have higher
accessibility and will thus be tracked with more implicit REs, while
others that have lower accessibility will be tracked with more explicit
REs Also based on previous research, we hypothesize that late signers
will become more over-redundant in their reference tracking compared to
native signers. This would translate to a preference by late signers for
nominals over null marking especially in maintenance and re-introduction
contexts. In contrast, we would expect that native signers use more null
marking to track referents.

@@

We tested 29 deaf adult signers, 15 with exposure to TİD from birth
(i.e., native signers), 14 with exposure between ages 3 and 17 (i.e.,
late signers) (\emph{M}\textsubscript{Age of acquisition} = 7.7 years,
\emph{SD} = 3.3 years). The criterion to be classified as a late signer
was to have hearing and non-signing caregivers. All the signers
satisfying this criterion were regarded as late signers
\footnote{Note that we are essentially comparing deaf of deaf and deaf of hearing adult signers. Although we did not set a cut-off point for nativeness, all the deaf of hearing participants indicated that they learned TİD after age 3. In order to capture in-group variation among late signers, we run an additional supplementary analysis}.
All participants indicated that their preferred language of
communication was TİD which they used for most contexts. Detailed
information about the linguistic and educational background of the
participants can be found in \ref{section:a}.

@@

Comparably, late learners (8 males, 6 females) were all born into
hearing families and started learning TİD later in life after age 3,
usually with the start of formal education. The age of sign language
acquisition for late signers ranged between years 3-17 (\emph{M} = 8.7
years, \emph{SD} = 3.5 years). The parents of all participating late
signers were hearing, and the reported home languages used in childhood
mainly consisted of Turkish and home sign. The average age of late
signers was 34.6 years (\emph{SD} = 8 years) and ranged between 24 and
50. The derived mean number of years for TİD use following its
acquisition was comparable to native signers with 25.9 years (\emph{SD}
= 7.9 years). Except for two signers who became deaf between the ages of
4 and 7, all the other late signers were pre-lingually deaf. All late
signers completed middle school education and the average year of
schooling was 12.3 years (\emph{SD} = 2.8 years). All late participants
attended at least one deaf school. While 8 participants attended two
deaf schools (either primary and middle school, or middle and high
school), 3 participants reported attend three deaf schools. Using the
same scale, late signers self-rated their abilities in TİD and (spoken)
Turkish. Both of their receptive and productive TİD abilities rated on
average 4.9 points (\emph{SD} = 0.3 points). Late signers' mean rating
for their reading abilities in Turkish was 3.5 points (\emph{SD} = 0.7
points) while their writing and speaking abilities rated on average 3.4
(\emph{SD} = 0.5 points) and 3.1 points (\emph{SD} = 0.7 points),
respectively.

@@

\begin{tabular}{p{3.2cm}p{3.8cm}p{3.8cm}}
\toprule
Tier & \multicolumn{1}{c}{Tag} & \multicolumn{1}{c}{Description}\\
\midrule
Clause Boundary &  & The boundary between clauses indicated by the presence of predicates and/or certain prosodic cues (e.g., head nod)\\
Discourse Status & Introduced & A first mention of an entity in a given discourse regardless of any syntactic role\\
 & Maintained & A subject referent continued across two or more immediate clauses\\
 & Reintroduced & A subject referent discontinued across at least one clause\\
REAT Type & NOM: Bare Noun & A non-modified noun phrase (e.g., CAT, MOUSE)\\
 & NOM: Modified Noun & A noun phrase modified by an index sign or an adjective\\
 & NOM: Fingerspelled Noun & Borrowed nouns from Turkish spelled by the manual alphabet (e.g., K-E-D-İ ‘cat’)\\
 & PRO: Pronominal IX & An index sign used in isolation to refer to an entity in discourse\\
 & CL: Whole Entity Classifier & A manual construction that refers to the entirety of an entity\\
 & CL: Body Part Classifier & A manual construction that refers to a part of an entity\\
 & CL: Handling Classifier & A manual construction that refers to the handling of an item\\
 & CL: Extension Classifier & A manual construction that denotes the size and shape of an entity\\
 & NULL: Constructed Action & A multifunctional construction where the signers take on the role of a referent and perform their actions\\
 & NULL: Plain Verb & A non-agreeing verb (e.g., LOVE)\\
 & NULL: Agreement Verb & A verbal construction the beginning and ending of which agree with person and number\\
\bottomrule
\addlinespace
\end{tabular}

@@

Participants sat in front of a laptop computer which had the stimulus
stories. The instructions were all given by a deaf research assistant
who was present in all sessions and ensured that the task requirements
were clear. The informed consent form was translated into TİD and
explained to the participants by the same assistant before data
recording. Participants filled in the consent form and a background
questionnaire form before the onset of the trials. Following this step,
signers were seated across a hearing researcher who is an advanced late
learner of TİD. We instructed the participants to narrate as they would
to a deaf friend. All trials were recorded by a digital camera placed in
front of the signers. The participants were instructed to watch the
narrative clip as many times as they needed, and narrate the content of
each clip either to the experimenter or the recording digital
camera\footnote{Only 4 participants (2 native, 2 late) narrated to the experimenter whereas the remaining e did not make it obligatory for participants to either narrate to the digital camera or to the experimenter for few reasons. First, a native or native-like deaf experimenter was not always available, and a hearing late signer of TİD attended some sessions as the experimenter. Therefore, most signers preferred to narrate to the camera. Furhermore, some signers did not feel comfortable narrating to the camera and, as a result, narrated to the deaf research assistant, if available. We ran an additional model excluding the 4 signers who narrated to the experimenter and compared it with the original model including these signers (see \ref{section:c} The results imply that there is not much of a difference.}.

@@

We collected 4142 data points as a result of the experiment. Each data
point is a referent tracked by the participant. Each of the REs is
classified in terms of its Discourse Status (Introduction, Maintenance,
Re-introduction), its Type (Nominal, Pronominal, Classifer,
Null)\footnote{In our statistical analyses, we will not consider the subtypes of REs but rather group them by the four proposed main types (Nominal, Pronominal, Classifer Null). However, we allude to the numerical distribution of the subtypes in this and following sections},
and Participant Group (Native,
Late)\footnote{Similarly, we treat late signers as a single group in our analyses and disregard the differing age of acquisition intervals among the late signers (0-3, 4-7, 8-12, 13-17). Treating age of acquisition as a continous variable would not be informative due to the small sample size in each interval.}.

@@

Among the null markers, CA was most frequently employed, and it was
followed by plain verbs and agreement verbs (Table \ref{tab:my-table5}).
Overall, signers used very few null markings for the introduction of a
referent and the majority of the existing ones were CA. Additionally,
signers used CA more for referent re-introduction than maintenance. We
observed a similar instances of agreement verbs for maintained and
re-introduced contexts. In contrast, plain verbs were favored more in
the maintenance setting than re-introduction.

@@

\begin{figure}
\centering
\includegraphics{manuscript_v2_files/figure-latex/fig:count-plot-1.pdf}
\caption{Referent counts by Discourse Status (Introduction, Maintanence,
Re-introduction), RE Type (Nominal, Pronominal, Classifier, Null), and
Participant Group (Native, Late)}
\end{figure}

>>>>

Sign Languages are natural languages of the Deaf communities all around
the world. They are independent languages, which do not derive from
majority spoken language(s) but emerge naturally when deaf people form a
community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has
been the officially recognized language of the Deaf community in Turkey
since 2005 (see Nuhbalaoglu, 2018 for a brief review of the historical
development of the language). Deaf or hearing children acquiring a SL
within a family where there is one or more deaf caregivers display the
same developmental milestones as hearing children acquiring a spoken
language in a hearing family (Chamberlain et al., 1999; Meier and
Newport, 1990). However, reports indicate that over 90\% of all deaf
individuals are born into hearing families who do not know how to sign
(i.e., those with language deprivation) (Woll, 2013). This means that
the auditory modality used in the household is not accessible to many
deaf children. That is why the age of acquisition of a first functional
language varies greatly among deaf children, with many of whom being
deprived of their first languages typically until after age 7 when they
start formal education in a deaf school. First language deprivation or
delay has been found to have negative effects on certain cognitive and
linguistic measures among late signing children. To name a few, delayed
exposure to a first functional language is reported to negatively affect
deaf children and adults' morphosyntactic production and processing
(Boudreault and Mayberry, 2006; Cheng and Mayberry, 2021; Cormier et
al., 2012; Karadöller et al., 2017; Kayabaşı et al., 2020; Newport,
1990) and narrative abilities (e.g., Becker, 2009). In our overall
research, we are concerned with the effects of this late exposure in the
ultimate adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy (Gundel et al., 1993).
According to this model, given or familiar information is expressed in
more attenuated forms such as pronominal use and null marking since the
mental status of the expressed referent is already at the center of
attention, as indicated by the present contextual information (Chafe,
1976). The Givenness Hierarchy is akin to other previously proposed
models of reference control within the topicality and saliency paradigm
such as the Accessibility Hierarchy (Ariel, 1990), Topic Continuity
(Givon, 1983), and Familiarity Scale (Prince, 1981). Gundel et al.
(1993) argued that the distribution of forms is not random but abides by
the Gricean Maxim of Quantity (Grice, 1975). That is to say, individuals
must make the most informative and succinct choice possible to retain
economy of form. When multiple expressions are available for the same
referent as in an utterance like (\ref{ex:1}), phonetically simplest
form is selected with the least redundant information based on an
economy principle. In the utterance, REs for ``Eddie'' are given in
bold. All three REs (nominal, pronominal, and null marking) refer to the
same entity; however, the quantity of marking decreases as the
referent's accessibility increases. On this ground, for highly
accessible referents (i.e., those that are highly active in the mind),
linguistic forms with the least phonetic content are selected for the
least amount of processing effort.

@@

Commonly in both spoken and sign languages nouns and noun phrases are
used as REs. Spoken languages use pronouns, determines and
demonstratives where sign languages use a pointing sign, usually glossed
as IX, which assumes similar functions (Lillo-Martin and Klima, 1990).
In addition, the pointing sign is used to referentially anchor a
referent to a location in signing space. For instance, when the signer
introduces \textsc{elvan} for the first time, s/he can assign her to the
right. This referential locus can later be used by agreement verbs and
classifier predicates (more below). For instance, in (\ref{ex:2a}) there
are two referents, \textsc{elvan}, who is referentially-assigned to the
right side of the signer, represented with the ``A'' subscript and
\textsc{demet}, who is referentially-assigned to the left side of the
signer, represented with the ``B'' subscript. When the signer makes a
contrast in (\ref{ex:2b}), as a pronominal use, she points to the right
side to refer to ELVAN and to the left to refer to \textsc{demet}. We
schematically represent the referential locus assigned to the referents
in (\ref{ex:3}).

@@

Grammatical gender as in Romance languages or noun classes as in Bantu
languages are some reference tracking tools that are familiar from
spoken languages. A somehow less known reference-tracking tool in spoken
languages is \emph{verbal classifiers}, which morphologically attach to
the verb and often cross-reference a core argument in a clause
(Aikhenvald, 2017). Classifiers denote sets of objects based on semantic
criteria such as size, shape, body-part, texture and sometimes function.
Usually, the verbal classifier and the cross-referenced noun co-occur in
the same clause when the argument is introduced into discourse. The
classifier can track the argument on its own on subsequent occurrences.
In sign languages, a verbal classifier morpheme is encoded by a
\emph{handshape} that is articulated phonologically co-temporally with a
movement, which encodes the verbal meaning (Benedicto and Brentari,
2004; Supalla, 1990). These two infused morphemes form a complex
predicate. Some classifier handshapes and the set of objects that they
denote are presented below in (\ref{ex:4}).

@@

In (\ref{ex:5}), in the first clause, a referential locus, i.e.~locus-A,
is assigned to the locative source argument \textsc{cage}. In the second
clause, the core argument \textsc{bird} is cross-referenced with the
classifier handshape, i.e.~small-legged animal. It flies from the
referential locus of the \textsc{cage} (locus-a) to the referential
locus of the goal \textsc{tree} (locus-B).

@@

    \ex \textsc{tree} assigned to the referential locus-b. The source and goal tracked on the predicates \textsc{fly} and \textsc{land}. \\ \\
    \textsc{bird fly[manner]-a tree-b land cl-b}\\
    ‘The bird flies (from the cage assigned to the left) to a tree (assigned to the right).’
    \end{xlist}
    \end{exe}
    \hfill (Adapted from Gökgöz and Sevgi 2020:318)

@@

Classifier predicates are highly complex morphological constructions
which are very common in sign languages (Supalla, 1986). These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape (Talmy, 2003). As in most sign languages hitherto studied,
classifier constructions are also commonly found in TİD (see Dikyuva et
al., 2015 for a review). Following the classification of
Engberg-Pedersen (1993), we distinguish between four types of classifier
constructions. Whole entity classifiers (WECL) represent the entirety of
any referent whereas body part classifiers (BPCL) only reflect a
specific part of an animate entity (e.g., limbs, head, tongue, teeth,
and so forth). Handling classifiers (HCL) occur with transitive verbs
and depict how an object is handled or manipulated by an agent. On this
ground, an HCL\footnote{As discussed before, an HCL is transitive in
  nature and depicts the handling action performed by ananimate agent.
  This is why most research tend to consider HCLs as an extension to
  constructed action (Cormier et al., 2013; Dudis, 2004).} depicts the
manner of the referring action while the former two constructions encode
the location of the entity as well as its motion in space. Extension
classifiers (ExtCL), on the other hand, denote the size and shape of an
entity by following a trajectory to trace its shape. We exemplify these
constructions in (\ref{ex:8}).

@@

Like in first language development, narrative skills have been
extensively studied among second language learners (see Bel et al.,
2015; Frederiksen and Mayberry, 2019). Comparing 11 deaf adult signers
and 13 hearing advanced learners of Catalan Sign Language (Llengua de
Signes Catalana - LSC), Bel et al. (2015) found that hearing signers
resort to more overt forms (nominals and pronouns) than native signers
to avoid potential ambiguities among referents. Nevertheless, their
results did not reach statistical significance for the interaction of
discourse status and acquisition group. A similar line of research came
from Frederiksen and Mayberry (2019) who found a statistically
significant difference between both groups, that is native signers and
hearing L2 learners of ASL, only for classifier use especially in
re-introduced contexts. L2 learners in the study used more nominals and
fewer instances of null marking in number, indicative of a limited
effect of overexplicitness among second language learners. However,
these studies treated late and native deaf signers as a single group.
That is, the control group consisted of both native signers who had one
or more deaf parents and also native-like (Cormier et al., 2012) late
deaf signers who had hearing parents and their age of acquisition
varied. Given the preliminary evidence that topic continuity and
salience marking is also difficult for second language learners
regardless of modality, it has been suggested under the Interface
Hypothesis that the syntax-discourse interface is vulnerable and
especially challenging to acquire (Sorace, 2011) for second language
learners. Similarly, Sorace and Serratrice (2009) discuss that this
interface entails much more than morphosyntactic knowledge but also
includes cognitive and pragmatic resources that help provide discourse
coherence. They propose that the second language discourse is
over-redundant. This means that second language learners produce more
overt pronouns or fuller forms of language for highly accessible
referents possibly as a compensatory strategy in an attempt to reduce
the increasing load on cognitive processing.

Still, only very few researchers examined whether linguistically
deprived signers (i.e., deaf signers with delayed exposure to a sign
language) also display redundancy in their reference tracking. Becker
(2009) examined 6 DGS signing deaf children's (3 native and 3 late)
narrative abilities. She reported that late signing children generated
more explicit forms such as nominals when referring to the characters in
the story compared to native learners, concluding that first language
deprivation results in an inadequate representation of the addressee and
less efficient use of morphosyntax. Comparably, Cormier et al. (2013)
investigated the use of CA in BSL among 15 deaf native, early, and late
adult signers (5 participants in each group) and found that late signers
were redundant in topic continuity. In other words, the late signers
produced more explicit forms like nominals, and fewer instances of null
marking when maintaining previous referents. Most relevantly to the
present study, Gür (2018) studied the development of narrative skills
among 10 native and 10 late deaf children and deaf adult signers of TİD.
She found that native-signing children and adults had more instances of
explicit referent introduction, set the scene more frequently, and
employed CA in an appropriate manner more than corresponding
late-signing children and adults. Although this study presented
preliminary findings on age of acquisition effects on narrative
competence, it does not investigate how native and late signers employ
different linguistic tools based on the accessibility of the referent.
Therefore, to our knowledge, there is a lack of research directly
examining the effects of first language deprivation on deaf adult
signers' selection of REs in signed narratives.

@@

Overall, considering what has been suggested for the distribution of REs
in signed narratives, we expect more explicit forms like nominals for
referent introduction and and less explicit forms such as null marking,
pronominal and/or classifier use for referent maintainance. For the
re-introduction contexts, where referents which are discontinued across
one or more clauses are brought back into the discourse, we expect a
less clear picture. Some re-introduced referents will have higher
accessibility and will thus be tracked with more implicit REs, while
others that have lower accessibility will be tracked with more explicit
REs. Also, based on previous research, we hypothesize that late signers
will be more over-redundant in their reference tracking compared to
native signers. This would translate to a preference by late signers for
nominals over null marking especially in maintenance and re-introduction
contexts. In contrast, we would expect that native signers use more null
marking to track referents.

@@

We tested 29 deaf adult signers, 15 with exposure to TİD from birth
(i.e., native signers), 14 with exposure between ages 3 and 17 (i.e.,
late signers) (\emph{M}\textsubscript{Age of acquisition} = 7.7 years,
\emph{SD} = 3.3 years). The criterion to be classified as a late signer
was to have hearing and non-signing caregivers. All the signers
satisfying this criterion were regarded as late signers
\footnote{Note that we are essentially comparing deaf of deaf and deaf of hearing adult signers. Although we did not set a cut-off point for nativeness, all the deaf of hearing participants indicated that they learned TİD after age 3. In order to capture in-group variation among late signers, we run an additional supplementary analysis in which we observed very wide credible intervals due to the small group size, rendering the results uninformative. For the analysis, see \ref{section:d}}.
All participants indicated that their preferred language of
communication was TİD which they used for most contexts. Detailed
information about the linguistic and educational background of the
participants can be found in \ref{section:a}.

@@

Comparably, late learners (8 males, 6 females) were all born into
hearing families and started learning TİD later in life after age 3,
usually with the start of formal education. The age of sign language
acquisition for late signers ranged between years 3-17 (\emph{M} = 8.7
years, \emph{SD} = 3.5 years). The parents of all participating late
signers were hearing, and the reported home languages used in childhood
mainly consisted of Turkish and home sign. The average age of late
signers was 34.6 years (\emph{SD} = 8 years) and ranged between 24 and
50. The derived mean number of years for TİD use following its
acquisition was comparable to native signers with 25.9 years (\emph{SD}
= 7.9 years). Except for two signers who became deaf between the ages of
4 and 7, all the other late signers were pre-lingually deaf. All late
signers completed at least middle school education and the year of
schooling ranged between 8 and 16 years, and the average was 12.3 years
(\emph{SD} = 2.8 years). All late participants attended at least one
deaf school. While 8 participants attended two deaf schools (either
primary and middle school, or middle and high school), 3 participants
reported to attend three deaf schools. Using the same scale, late
signers self-rated their abilities in TİD and Turkish. Both of their
receptive and productive TİD abilities rated on average 4.9 points
(\emph{SD} = 0.3 points). Late signers' mean rating for their reading
abilities in Turkish was 3.5 points (\emph{SD} = 0.7 points) while their
writing and speaking abilities rated on average 3.4 (\emph{SD} = 0.5
points) and 3.1 points (\emph{SD} = 0.7 points), respectively.

@@

\begin{tabular}{p{3.2cm}p{3.8cm}p{3.8cm}}
\toprule
Tier & \multicolumn{1}{c}{Tag} & \multicolumn{1}{c}{Description}\\
\midrule
Clause Boundary &  & The boundary between clauses indicated by the presence of predicates and/or certain prosodic cues (e.g., head nod)\\
Discourse Status & Introduced & A first mention of an entity in a given discourse regardless of any syntactic role\\
 & Maintained & A subject referent continued across two or more immediate clauses\\
 & Reintroduced & A subject referent discontinued across at least one clause\\
RE Type & NOM: Bare Noun & A non-modified noun phrase (e.g., CAT, MOUSE)\\
 & NOM: Modified Noun & A noun phrase modified by an index sign or an adjective\\
 & NOM: Fingerspelled Noun & Borrowed nouns from Turkish spelled by the manual alphabet (e.g., K-E-D-İ ‘cat’)\\
 & PRO: Pronominal IX & An index sign used in isolation to refer to an entity in discourse\\
 & CL: Whole Entity Classifier & A manual construction that refers to the entirety of an entity\\
 & CL: Body Part Classifier & A manual construction that refers to a part of an entity\\
 & CL: Handling Classifier & A manual construction that refers to the handling of an item\\
 & CL: Extension Classifier & A manual construction that denotes the size and shape of an entity\\
 & NULL: Constructed Action & A multifunctional construction where the signers take on the role of a referent and perform their actions\\
 & NULL: Plain Verb & A non-agreeing verb (e.g., LOVE)\\
 & NULL: Agreement Verb & A verbal construction the beginning and ending of which agree with person and number\\
\bottomrule
\addlinespace
\end{tabular}

@@

Participants sat in front of a laptop computer which had the stimulus
stories. The instructions were all given by a deaf research assistant
who was present in all sessions and ensured that the task requirements
were clear. The informed consent form was translated into TİD and
explained to the participants by the same assistant before data
recording. Participants filled in the consent form and a background
questionnaire form before the onset of the trials. Following this step,
signers were seated across a hearing researcher who is an advanced late
learner of TİD. We instructed the participants to narrate as they would
to a deaf friend. All trials were recorded by a digital camera placed in
front of the signers. The participants were instructed to watch the
narrative clip as many times as they needed, and narrate the content of
each clip either to the experimenter or the recording digital
camera\footnote{Only 4 participants (2 native, 2 late) narrated to the experimenter whereas the remaining we did not make it obligatory for participants to either narrate to the digital camera or to the experimenter for few reasons. First, a native or native-like deaf experimenter was not always available, and a hearing late signer of TİD attended some sessions as the experimenter. Therefore, most signers preferred to narrate to the camera. Furhermore, some signers did not feel comfortable narrating to the camera and, as a result, narrated to the deaf research assistant, if available. We ran an additional model excluding the 4 signers who narrated to the experimenter and compared it with the original model including these signers (see \ref{section:c}). The results imply that there is not much of a difference.}.

@@

We collected 4142 data points as a result of the experiment. Each data
point is a referent tracked by the participant. Each of the REs is
classified in terms of its Discourse Status (Introduction, Maintenance,
Re-introduction), its Type (Nominal, Pronominal, Classifer,
Null)\footnote{In our statistical analyses, we will not consider the subtypes of REs but rather group them by the four proposed main types (Nominal, Pronominal, Classifer, and Null). However, we allude to the numerical distribution of the subtypes in this and following sections},
and Participant Group (Native,
Late)\footnote{Similarly, we treat late signers as a single group in our analyses and disregard the differing age of acquisition intervals among the late signers (0-3, 4-7, 8-12, 13-17).}.

@@

Among the null markers, CA was most frequently employed, and it was
followed by plain verbs and agreement verbs (Table \ref{tab:my-table5}).
Overall, signers used very few null markings for the introduction of a
referent and the majority of the existing ones were CA. Additionally,
signers used CA more for referent re-introduction than maintenance. We
observed similar instances of agreement verbs for maintained and
re-introduced contexts. In contrast, plain verbs were favored more in
the maintenance setting than re-introduction.

@@

\begin{figure}
\centering
\includegraphics{manuscript_v2_files/figure-latex/count-plot-1.pdf}
\caption{Referent counts by Discourse Status (Introduction, Maintanence,
Re-introduction), RE Type (Nominal, Pronominal, Classifier, Null), and
Participant Group (Native, Late)}
\end{figure}

For more inference on the data and having pairwise comparisons, we fit a
regression model to RE Type using discourse status and acquisition group
as predictors. We used the brms package (Bürkner, 2018) in R, with a
categorical distribution on the responses and sum contrasts for the
predictors Discourse Status (Introduction, Maintenance, Re-introduction)
and Participant Group (Native, Late). As the overall number of produced
pronouns was very little (40 data points), we excluded pronominal use
from the following analyses. For the current analysis, we did not define
priors ourselves and used the preset uninformed priors.
@ref\{fig:model-gram\}\footnote{We ran another model where we used the
  Nativeness predictor with 3 levels 0-3:Native, 4-7:Late1, and
  7-above:Late2. However, our model did not converge properly and the
  credible intervals were very wide spread, yielding the results
  uninformative. We provide the model results as a plot in
  \ref{section:d}.} shows the posterior probability distributions where
the dot is the median estimate, the thick line is the 50\% credible
intervals, and the thin line is the 95\% credible intervals. There are 3
responses we are considering in this model. They are; Classifier,
Nominative, and Null. Classifier response is the level that other
responses Nominative and Null are compared against.

\begin{verbatim}
## Warning: Ignoring unknown aesthetics: height
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_v2_files/figure-latex/model-gram-1.pdf}
\caption{Regression model plot for the RE Type (Nominal, Null) with the
predictors Discourse Status (Introduction, Maintenance, Re-introduction)
and Participant Group (Native, Late)}
\end{figure}
